Open Data, Open Source and Open Standards in chemistry: \\ The Blue Obelisk five years on

### 2011-06-15 paper.tex
Minor corrections

\author{
  Noel M O'Boyle\correspondingauthor$^1$%
  \email{Noel M O'Boyle\correspondingauthor - baoilleach@gmail.com}
  \and
  Samuel E Adams$^2$%
  \email{Samuel E Adams - sea36@cam.ac.uk}
  \and
  Jonathan Alvarsson$^3$%
  \email{Jonathan Alvarsson - jonathan.alvarsson@farmbio.uu.se}
  \and
  Richard L Apodaca$^{4}$%
  \email{Richard L Apodaca - rapodaca@metamolecular.com}
  \and
  Jean-Claude Bradley$^5$%
  \email{Jean-Claude Bradley - bradlejc@drexel.edu}
  \and
  Igor V Filippov$^6$%
  \email{Igor V Filippov - igorf@helix.nih.gov}
  \and
  Rajarshi Guha$^6$%
  \email{Rajarshi Guha - guhar@mail.nih.gov}
  \and
  Robert M Hanson$^7$%
  \email{Robert M Hanson - hansonr@stolaf.edu}
  \and
  Marcus D Hanwell$^{8}$%
  \email{Marcus D Hanwell - marcus.hanwell@kitware.com}
  \and
  Geoffrey R Hutchison$^9$%
  \email{Geoffrey R Hutchison - geoffh@pitt.edu}
  \and
  Andrew SID Lang$^{10}$%
  \email{Andrew SID Lang - alang@oru.edu}
  \and
  Karol M Langner$^{11}$%
  \email{Karol M Langner - langnerkm@chem.leidenuniv.nl}
  \and
  Daniel M Lowe$^2$%
  \email{Daniel Lowe - dl387@cam.ac.uk}
  \and
  Dmitry Pavlov$^{12}$%
  \email{Dmitry Pavlov - dpavlov@ggasoftware.com}
  \and
  Ola Spjuth$^3$%
  \email{Ola Spjuth - ola.spjuth@farmbio.uu.se}
  \and
  Christoph Steinbeck$^{13}$%
  \email{Christoph Steinbeck - steinbeck@ebi.ac.uk}
  \and
  Adam L Tenderholt$^{14}$%
  \email{Adam L Tenderholt - adamlt82@u.washington.edu}
  \and
  Kevin J Theisen$^{15}$%
  \email{Kevin J Theisen - kevin@ichemlabs.com}
  \and
  Egon L Willighagen$^{16}$%
  \email{Egon L Willighagen - egon.willighagen@ki.se}
  \and
  Peter Murray-Rust$^2$%
  \email{Peter Murray-Rust - pm286@cam.ac.uk}
}

@@

\address{%
    \iid(1)Analytical and Biological Chemistry Research Facility, Cavanagh Pharmacy Building, University College Cork, College Road, Cork, Co. Cork, Ireland \\
    \iid(2)Unilever Centre for Molecular Sciences Informatics, Department of Chemistry, University of Cambridge, Lensfield Road, CB2 1EW, UK \\
    \iid(3)Department of Pharmaceutical Biosciences, Uppsala University, Box 591, 751 24 Uppsala, Sweden \\
    \iid(4)Metamolecular, LLC, 8070 La Jolla Shores Drive \#464, La Jolla, CA 92037, USA \\
    \iid(5)Department of Chemistry, Drexel University, 32nd and Chestnut streets, Philadelphia, PA 19104, USA \\
    \iid(6)NIH Center for Translational Therapeutics, 9800 Medical Center Drive, Rockville, MD 20878, USA \\
    \iid(7)St. Olaf College, 1520 St. Olaf Ave., Northfield, MN 55057, USA \\
    \iid(8)Kitware, Inc., 28 Corporate Drive, Clifton Park, NY 12065, USA \\
    \iid(9)Department of Chemistry, University of Pittsburgh, 219 Parkman Avenue, Pittsburgh, PA 15260, USA \\
    \iid(10)Department of Engineering, Computer Science, Physics, and Mathematics, Oral Roberts University, 7777 S. Lewis Ave. Tulsa, OK 74171, USA \\
    \iid(11)Leiden Institute of Chemistry, Universiteit Leiden, Einsteinweg 55, 2333 CC Leiden, The Netherlands \\
    \iid(12)GGA Software Services LLC, 41 Nab. Chernoi rechki 194342, Saint Petersburg, Russia \\
    \iid(13)Cheminformatics and Metabolism Team, European Bioinformatics Institute (EBI), Wellcome Trust Genome Campus, Hinxton, Cambridge, UK \\
    \iid(14)Department of Chemistry, University of Washington, Seattle, WA 98195, USA \\
    \iid(15)iChemLabs, 200 Centennial Ave., Suite 200, Piscataway, NJ 08854, USA \\
    \iid(16)Division of Molecular Toxicology, Institute of Environmental Medicine, %
        Nobels vaeg 13, Karolinska Institutet, 171 77 Stockholm, Sweden 
}%

@@

        \paragraph*{Conclusions:} We show that the Blue Obelisk has been very successful
		in bringing together researchers and developers with common interests
		in ODOSOS, leading to development of many useful resources freely
		available to the chemistry community. But how best to engage with the
		wider chemistry community outside of the Blue Obelisk remains an open
		question.
\end{abstract}

@@

However, Blue Obelisk software and data is used outside
cheminformatics and certainly in the five main areas that, for
example, Chemical Markup Language (CML) \cite{murray-rust_chemical_1999} supports:

@@

While desktop software has composed the majority of scientific tools
since the computer was introduced, the internet continues to change
how applications and content are distributed and presented. The web
presents new opportunities for scientists as it is
an open and free medium to distribute scientific knowledge, ideas and
education. Web applications are software that runs within the browser,
typically implemented in Java or JavaScript.
Recently, a new version of the HTML
specification, HTML5, defines a well-developed framework
for creating native web applications in JavaScript and opens up
new possibilities for visualising chemical data.

@@

\subsection*{QSAR-ML}
The field of QSAR has long been hampered by the lack of open
standards, which makes it difficult to share and reproduce descriptor
calculations and analyses. QSAR-ML was recently proposed as an open
standard for exchanging QSAR datasets~\cite{Spjuth:2010uq}. A dataset
in QSAR-ML includes the chemical structures (preferably described in
CML) with InChI to protect integrity, chemical descriptors linked to
the Blue Obelisk Descriptor Ontology~\cite{bodo}, response values,
units, and versioned descriptor implementations to allow for
integrating several descriptor software in the same calculation.
Hence, a dataset described in QSAR-ML is completely reproducible. To
allow for easy setup of QSAR-ML compliant datasets, a plugin for
Bioclipse was created with a graphical interface for
setting up QSAR datasets and performing calculations. Descriptor
implementations are available from the CDK and JOELib, as well
as via remote web services such as XMPP~\cite{Wagener:2009uq}.

@@

We have shown that the Blue Obelisk has been very successful
in bringing together researchers and developers with common interests
in ODOSOS, leading to development of many useful resources freely
available to the chemistry community. But how best to engage with the
wider chemistry community outside of the Blue Obelisk remains an open
question. If the Blue Obelisk is truly to make an impact,
then an attempt must be made to reach beyond the subscribers to the
Blue Obelisk mailing list and blogs of members.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Acknowledgements}
  \ifthenelse{\boolean{publ}}{\small}{}
  NMOB is supported by a Health Research Board Career Development
Fellowship (PD/2009/13).

>>>>

\author{
  Noel M O'Boyle\correspondingauthor$^1$%
  \email{Noel M O'Boyle\correspondingauthor - baoilleach@gmail.com}
  \and
  Samuel E Adams$^2$%
  \email{Samuel E Adams - sea36@cam.ac.uk}
  \and
  Jonathan Alvarsson$^3$%
  \email{Jonathan Alvarsson - jonathan.alvarsson@farmbio.uu.se}
  \and
  Richard L Apodaca$^{4}$%
  \email{Richard L Apodaca - rapodaca@metamolecular.com}
  \and
  Jean-Claude Bradley$^5$%
  \email{Jean-Claude Bradley - bradlejc@drexel.edu}
  \and
  Igor V Filippov$^6$%
  \email{Igor V Filippov - igorf@helix.nih.gov}
  \and
  Rajarshi Guha$^7$%
  \email{Rajarshi Guha - guhar@mail.nih.gov}
  \and
  Robert M Hanson$^8$%
  \email{Robert M Hanson - hansonr@stolaf.edu}
  \and
  Marcus D Hanwell$^{9}$%
  \email{Marcus D Hanwell - marcus.hanwell@kitware.com}
  \and
  Geoffrey R Hutchison$^{10}$%
  \email{Geoffrey R Hutchison - geoffh@pitt.edu}
  \and
  Andrew SID Lang$^{11}$%
  \email{Andrew SID Lang - alang@oru.edu}
  \and
  Karol M Langner$^{12}$%
  \email{Karol M Langner - langnerkm@chem.leidenuniv.nl}
  \and
  Daniel M Lowe$^2$%
  \email{Daniel Lowe - dl387@cam.ac.uk}
  \and
  Dmitry Pavlov$^{13}$%
  \email{Dmitry Pavlov - dpavlov@ggasoftware.com}
  \and
  Ola Spjuth$^3$%
  \email{Ola Spjuth - ola.spjuth@farmbio.uu.se}
  \and
  Christoph Steinbeck$^{14}$%
  \email{Christoph Steinbeck - steinbeck@ebi.ac.uk}
  \and
  Adam L Tenderholt$^{15}$%
  \email{Adam L Tenderholt - adamlt82@u.washington.edu}
  \and
  Kevin J Theisen$^{16}$%
  \email{Kevin J Theisen - kevin@ichemlabs.com}
  \and
  Egon L Willighagen$^{17}$%
  \email{Egon L Willighagen - egon.willighagen@ki.se}
  \and
  Peter Murray-Rust$^2$%
  \email{Peter Murray-Rust - pm286@cam.ac.uk}
}

@@

\address{%
    \iid(1)Analytical and Biological Chemistry Research Facility, Cavanagh Pharmacy Building, University College Cork, College Road, Cork, Co. Cork, Ireland \\
    \iid(2)Unilever Centre for Molecular Sciences Informatics, Department of Chemistry, University of Cambridge, Lensfield Road, CB2 1EW, UK \\
    \iid(3)Department of Pharmaceutical Biosciences, Uppsala University, Box 591, 751 24 Uppsala, Sweden \\
    \iid(4)Metamolecular, LLC, 8070 La Jolla Shores Drive \#464, La Jolla, CA 92037, USA \\
    \iid(5)Department of Chemistry, Drexel University, 32nd and Chestnut streets, Philadelphia, PA 19104, USA \\
    \iid(6)Chemical Biology Laboratory, Basic Research Program, SAIC-Frederick,
Inc., NCI-Frederick, Frederick, MD 21702, USA \\
    \iid(7)NIH Center for Translational Therapeutics, 9800 Medical Center Drive, Rockville, MD 20878, USA \\
    \iid(8)St. Olaf College, 1520 St. Olaf Ave., Northfield, MN 55057, USA \\
    \iid(9)Kitware, Inc., 28 Corporate Drive, Clifton Park, NY 12065, USA \\
    \iid(10)Department of Chemistry, University of Pittsburgh, 219 Parkman Avenue, Pittsburgh, PA 15260, USA \\
    \iid(11)Department of Engineering, Computer Science, Physics, and Mathematics, Oral Roberts University, 7777 S. Lewis Ave. Tulsa, OK 74171, USA \\
    \iid(12)Leiden Institute of Chemistry, Leiden University, Einsteinweg 55, 2333 CC Leiden, The Netherlands \\
    \iid(13)GGA Software Services LLC, 41 Nab. Chernoi rechki 194342, Saint Petersburg, Russia \\
    \iid(14)Cheminformatics and Metabolism Team, European Bioinformatics Institute (EBI), Wellcome Trust Genome Campus, Hinxton, Cambridge, UK \\
    \iid(15)Department of Chemistry, University of Washington, Seattle, WA 98195, USA \\
    \iid(16)iChemLabs, 200 Centennial Ave., Suite 200, Piscataway, NJ 08854, USA \\
    \iid(17)Division of Molecular Toxicology, Institute of Environmental Medicine, %
        Nobels vaeg 13, Karolinska Institutet, 171 77 Stockholm, Sweden 
}%

@@

        \paragraph*{Conclusions:} We show that the Blue Obelisk has been very successful
		in bringing together researchers and developers with common interests
		in ODOSOS, leading to development of many useful resources freely
		available to the chemistry community.
\end{abstract}

@@

However, Blue Obelisk software and data is also used outside
cheminformatics and certainly in the five main areas that, for
example, Chemical Markup Language (CML) \cite{murray-rust_chemical_1999} supports:

@@

While desktop software has composed the majority of scientific tools
since the computer was introduced, the internet continues to change
how applications and content are distributed and presented. The web
presents new opportunities for scientists as it is
an open and free medium to distribute scientific knowledge, ideas and
education. Web applications are software that runs within the browser,
typically implemented in Java or JavaScript.
Recently, a new version of the HTML
specification, HTML5, defined a well-developed framework
for creating native web applications in JavaScript and this opens up
new possibilities for visualising chemical data.

@@

\subsection*{QSAR-ML}
The field of QSAR has long been hampered by the lack of open
standards, which makes it difficult to share and reproduce descriptor
calculations and analyses. QSAR-ML was recently proposed as an open
standard for exchanging QSAR datasets~\cite{Spjuth:2010uq}. A dataset
in QSAR-ML includes the chemical structures (preferably described in
CML) with InChI to protect integrity, chemical descriptors linked to
the Blue Obelisk Descriptor Ontology~\cite{bodo}, response values,
units, and versioned descriptor implementations to allow
descriptors from different software to be integrated into the same calculation.
Hence, a dataset described in QSAR-ML is completely reproducible. To
allow for easy setup of QSAR-ML compliant datasets, a plugin for
Bioclipse was created with a graphical interface for
setting up QSAR datasets and performing calculations. Descriptor
implementations are available from the CDK and JOELib, as well
as via remote web services such as XMPP~\cite{Wagener:2009uq}.

@@

We have shown that the Blue Obelisk has been very successful
in bringing together researchers and developers with common interests
in ODOSOS, leading to development of many useful resources freely
available to the chemistry community. However, how best to engage with the
wider chemistry community outside of the Blue Obelisk remains an open
question. If the Blue Obelisk is truly to make an impact,
then an attempt must be made to reach beyond the subscribers to the
Blue Obelisk mailing list and blogs of members.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Acknowledgements}
  \ifthenelse{\boolean{publ}}{\small}{}
  NMOB is supported by a Health Research Board Career Development
Fellowship (PD/2009/13).
This project has been funded in whole or in part with federal funds from
the National Cancer Institute, National Institutes of Health, under
contract HHSN261200800001E. The content of this publication does not
necessarily reflect the views of the policies of the Department of
Health and Human Services, nor does mention of trade names, commercial
products, or organisations imply endorsement by the US Government.
